The Impact of Alternation

### 2017-05-05 draft.tex
ok

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is a combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
\rightarrow \psi$ for $\neg\phi \vee \psi$.

We denote by $\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$
and write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, we denote by $\phi[t/x]$ the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$. A formula is in positive normal form (PNF) if negation
occurs only within literals and in disjunctive normal form (DNF) if it
is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
\bigwedge_{j=1}^{M_i} \ell_{ij}$. The size of a formula $\phi$,
denoted $\len{\phi}$ is the number of symbols occuring in it.

@@

An \emph{alternating data automaton} (or, simply, and automaton) is a
tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, also called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is a formula describing the
  \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final} states, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
current values of the variables of $\A$. Formally, let $\vec{x}_k =
\set{x_k \mid x \in \vec{x}}$, $k\geq0$, be a time-stamped set of
variables. For an input even $a \in \Sigma$ and a formula $\phi$ we
write $\Delta^k(\phi,a)$ for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. The size of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.

Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$, the
\emph{run} of $\A$ over $w$ is the sequence of formulae
$\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$ and, for
all $i > 0$, we have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. We
slightly abuse notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for
the formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the
following. $\A$ \emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where
$\nu$ is the valuation that maps:\begin{inparaenum}[(1)]
\item each $x \in \vec{x}_0$ to $\mathbf{0}$, 
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$, and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$, denoted as $L(\A)$, is the set of words accepted
by $\A$.

@@

\begin{lemma}\label{lemma:closure}
  For any automata $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
  and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$ such that $Q_1
  \cap Q_2 = \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$,
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{rcl}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iff & 
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee
  \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) \\
  & \iff & \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots
  a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{rcll}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & 
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
  & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
  \end{array}\]
  Let $\nu,\nu' : Q \cup \bigcup_{i=0}^n\vec{x}_i$ be two valuations
  such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x) = \mathbf{0}$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x = \overline{x} + k$, for $k \in \nat$ and $x \teq 0$,
with the classical interpretation of function symbols on
integers. Since reachability of a control state is undecidable for
AVASS, the emptiness problem is undecidable for ADA.

Consequently, we give up on the termination guarantee and develop
semi-algorithmic methods that are supposed to be
both: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a)$, mapping each formula in
$\Form^+(Q,\vec{x})$ to a formula, also in $\Form^+(Q,\vec{x})$, that
defines the effect of firing a transition with input event $a$.

@@

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, in practice, the antichain based semi-algorithm
diverges in a large number of cases, since, in many interpretations,
there exist infinite antichains\footnote{For instance, when $\Data^\I
  = \nat$, an infinite antichain is $q \wedge x \teq 0, q \wedge x
  \teq 1, \ldots$}.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
  \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
invariant. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration (in which all data variables are assumed to be
$\mathbf{0}\in\Data^\I$), over-approximates the set of reachable
valuations and excludes those valuations that satisfying the
acceptance condition.

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A)
= \emptyset$ iff $\A$ has a safety invariant. 
\end{lemma}
\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
  define a function $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x}
    \mapsto \Data^\I}$ as: for each $\beta : Q \rightarrow \booli$,
  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
  u \in \Sigma^* ~.~ \beta\cup\nu \models
  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is indeed a safety
  invariant is straightforward. \qed}

@@

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can occur negated in some $\phi \in \preds$. We assume that $\preds$
is fixed for now, and refer to these formulae as \emph{predicates}.
Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
w.r.t. the predicates in $\preds$. The abstract version of the
post-image and acceptance formulae are defined as follows:
{\footnotesize\[
\abs{\Post{\A}}(\phi,\varepsilon) \equiv \phi 
\hspace*{2mm}
\abs{\Post{\A}}(\phi,ua) \equiv \abs{(\Post{\A}(\abs{\Post{\A}}(\phi,u),a))} 
\hspace*{2mm}
\abs{\Accept{\A}}(u) \equiv \abs{\Post{\A}}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
\]}
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \abs{\Post{\A}}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k>0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [1,k]$ such that
  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\abs{\Post{\A}}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k > 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [1,k]$ such that $\abs{\Post{\A}}(\iota,u) \eqiff
  \abs{\Post{\A}}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

Once a counterexample $u$ is discovered, there are two
possibilities. Either\begin{inparaenum}[(i)]
\item $\Accept{\A}(u)$ is satisfiable, in which case $L(\A) \neq \emptyset$, or
\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is called
  \emph{spurious}.
\end{inparaenum}
In the first case, our semi-algorithm stops and returns a witness for
non-emptiness, obtained from the satisfying valuation of
$\Accept{\A}(u)$ and in the second case, we must strenghten the
invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
that $\beta\cup\nu \models \abs{\Accept{\A}}(u)$. This strenghtening
is carried out by adding to $\preds$ predicates sufficient to exclude
the spurious counterexample.

>>>>

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is an existentially quantified combination
of atoms using disjunction $\vee$, conjunction $\wedge$ and negation
$\neg$ and we write $\phi \rightarrow \psi$ for $\neg\phi \vee \psi$.

We denote by $\fv{S}{\phi}$ the set of free (i.e.\ not under the scope
of an existential quantifier) variables of sort $S$ in $\phi$ and
write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$.
%
%% A formula is in positive normal form (PNF) if negation occurs only
%% within literals and in disjunctive normal form (DNF) if it is a
%% disjunction of conjunctions of literals $\bigvee_{i=1}^N
%% \bigwedge_{j=1}^{M_i} \ell_{ij}$.
%
The size of a formula $\phi$, denoted $\len{\phi}$ is the number of
symbols occuring in it.

@@

An \emph{alternating data automaton} (or, simply, and automaton) is a
tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, also called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is a formula describing the
  \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final} states, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, in each formula describing a transition rule,
$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
values of the variables of $\A$. Observe further, that the initial
values of the variables are left unconstrained, because the initial
configuration does not contain free data variables. The size of $\A$
is defined as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a time-stamped set of variables. For an input even $a \in \Sigma$ and
a formula $\phi$ we write $\Delta^k(\phi,a)$ for the formula obtained
from $\phi$ by simultaneously replacing each state $q \in
\fv{\Bool}{\phi}$ by the formula
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k > 0$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. We slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the following. We say that
$\A$ \emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where $\nu$ is
some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$, and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.

@@

\begin{lemma}\label{lemma:closure}
  For any automata $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
  and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$ such that $Q_1
  \cap Q_2 = \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$,
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{rcl}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iff & 
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee
  \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) \\
  & \iff & \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots
  a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{rcll}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & 
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
  & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be two valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x \geq 0$ and $x = \overline{x} + k$, for $k \in \zed$,
with the classical interpretation of the function symbols on
integers. Since reachability of a control state is undecidable for
AVASS, the emptiness problem is undecidable for ADA.

Consequently, we give up on the guarantee for termination and develop
semi-algorithmic methods that are supposed to be
both: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
effect of reading $a \in \Sigma$.

@@

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, in practice, the antichain based semi-algorithm
diverges in a large number of cases, since, in many interpretations,
there exist infinite antichains of formulae\footnote{For instance,
  when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq 0,
  q \wedge x \teq 1, \ldots$}.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
  \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it simply an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota$, in which the data variables are unconstrained
and over-approximates the set of reachable valuations (\ref{it1:inv}),
while excluding those valuations that satisfying the acceptance
condition (\ref{it2:inv}).

\begin{lemma}\label{lemma:safety-invariant}
  For any automaton $\A$, we have $L(\A)
= \emptyset$ iff $\A$ has a safety invariant. 
\end{lemma}
\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
  u \in \Sigma^* ~.~ \beta\cup\nu \models
  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
  straightforward. \qed}

@@

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can occur negated in some $\phi \in \preds$. We assume that $\preds$
is fixed for now, and refer to these formulae as \emph{predicates}.
Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
w.r.t. the predicates in $\preds$. The abstract version of the
post-image and acceptance condition are defined as follows:
{\footnotesize\[ \abs{\Post{\A}}(\phi,\varepsilon) \equiv \phi
\hspace*{2mm}
\abs{\Post{\A}}(\phi,ua) \equiv \abs{(\Post{\A}(\abs{\Post{\A}}(\phi,u),a))} 
\hspace*{2mm}
\abs{\Accept{\A}}(u) \equiv \abs{\Post{\A}}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)\]}
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \abs{\Post{\A}}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\abs{\Post{\A}}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\abs{\Post{\A}}(\iota,u) \eqiff
  \abs{\Post{\A}}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

Once a counterexample $u$ is discovered, there are two
possibilities. Either\begin{inparaenum}[(i)]
\item $\Accept{\A}(u)$ is satisfiable, in which case $L(\A) \neq \emptyset$, or
\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is called
  \emph{spurious}.
\end{inparaenum}
In the first case, our semi-algorithm stops and returns a witness for
non-emptiness, obtained from the satisfying valuation of
$\Accept{\A}(u)$ and in the second case, we must strenghten the
invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
that $\beta\cup\nu \models \abs{\Accept{\A}}(u)$. This strenghtening
is carried out by adding to $\preds$ several predicates that are
sufficient to exclude the spurious counterexample.
